Hidden Talents in Context: Multiverse analysis repository

### 2022-04-05 manuscript/manuscript.Rmd
small fixes at the proofs stage

\noindent Adversity-exposed youth tend to score lower on cognitive tests. However, the hidden talents approach proposes some abilities are enhanced by adversity, especially under ecologically relevant conditions. Two versions of an attention shifting and working memory updating task---one abstract, one ecological---were administered to `r txt1_sam$non_arb$n$total` youth (*M~age~* = `r txt1_sam$non_arb$age$mean`, *SD~age~* = `r txt1_sam$non_arb$age$sd`; `r (txt1_sam$non_arb$sex$females/txt1_sam$non_arb$n$total * 100) %>% round(2)`% female; `r ((txt1_sam$non_arb$n$total-txt1_sam$non_arb$race$non_white)/txt1_sam$non_arb$n$total * 100) %>% round(2)`% White). Measures of environmental unpredictability, violence, and poverty were collected to tested adversity × task version interactions. There were no interactions for attention shifting. For working memory updating, youth exposed to violence and poverty scored lower than their peers with abstract stimuli but equally well with ecological stimuli. These results are striking compared to contemporary developmental science, which often reports lowered performance among adversity-exposed youth.

@@

One way to test this question is to replace sterile testing materials with ecologically relevant content from the real world. For example, one study tested memory and reasoning among young adults who varied in their exposures to violence [@frankenhuis2020a]. They created two versions of a memory and reasoning task, one neutral (chronological age) and one more ecologically relevant (social dominance). They found that people with more *current* exposures to violence were equally, or more, accurate when memorizing social-dominance relationships than people with fewer exposures. Notably, such violence exposures were associated with less accurate memory for neutral (age) relationships. This pattern was specific to *current* exposures: people with more *childhood* exposures to violence memorized both content types less accurately. There were no associations between violence exposures and reasoning performance. Overall, this study suggests that ecologically relevant testing materials may in some conditions 'level the playing field' for adversity-exposed youth [@ellis2020; @frankenhuis2020; @rifkin-graboi2021; @vantassel-baska2018].

Although research testing performance in ecologically relevant contexts or situations and using ecologically relevant materials have generated innovative ideas and findings, it remains unclear whether or how these ideas and findings can be practically applied. For example, narrow abilities, such as memorizing social dominance hierarchies or accurately detecting anger in faces, seem to have limited practical value in educational contexts. In addition, although executive functions and working memory are related to many educational outcomes, it is unclear how schools and teachers might leverage skills and abilities that specifically depend on inducing uncertainty. That seems impractical, at best, and unethical, at worst.

@@

Environmental unpredictability is difficult to measure and different definitions exist alongside each other [@young2020]. Despite this, we aligned our measurement approach with previous literature in two ways. First, we measured caregiver and residential instability, which are both thought to be proxies to environmental unpredictability, using interview-based measures [@ellis2009]. Second, we measured the extent to which youth perceive their family context as unpredictable using a questionnaire [@young2018]. Our goal was to aggregate these two information sources to arrive at a best estimate of environmental unpredictability.

@@

Our models for attention shifting and working memory updating differed in one respect. Our sample included a small subset of siblings (*n* = `r txt6_sib$n`, `r txt6_sib$families` families). All sibling groups were pairs of two (`r txt6_sib$sizes$two` sibling pairs) except for `r english::as.english(txt6_sib$sizes$three)` group of three and `r english::as.english(txt6_sib$sizes$four)` group of four siblings. We computed intraclass correlations (ICCs) for sibling groups for attention-shifting and working memory updating scores. ICCs for attention shifting were essentially zero, but they were non-zero for working memory updating scores (*ICC~standard~* = `r txt6_sib$iccs$abs_update`, *ICC~ecological~* = `r txt6_sib$iccs$eco_update`). Thus, all models analyzing working memory updating included an additional random intercept for siblings. Non-siblings were given the same sibling-group code, so the model could compute an intercept for non-siblings.

@@

Table \@ref(tab:table1) displays bivariate associations between all composite measures of adversity, including components of each, and descriptive statistics. Our primary interests were the main effect of adversity and the interaction of adversity with task-version. Table \@ref(tab:table2) shows median effect sizes ($\beta$-coefficients), sample sizes, and bootstrapped *p*-values for attention-shifting and working memory updating analyses for the main effect of adversity (unpredictability, violence, and poverty) and the interaction term (task version × adversity). Table \@ref(tab:table3) reports median simple effects of task version across high and low levels of adversity and median simple effects of adversity across abstract and ecologically relevant task versions.

@@

Figures \@ref(fig:figure2) and \@ref(fig:figure3) visualize multiverse task version × adversity interaction results for attention shifting and working memory performance, including depictions of performance on abstract and ecological task versions across high ($+$ 1 *SD*) and low ($-$ 1 *SD*) adversity exposure (Figures \@ref(fig:figure2)a and \@ref(fig:figure3)a), p-curves associated with each interaction term (Figures \@ref(fig:figure2)b and \@ref(fig:figure3)b), sorted interaction $\beta$-coefficients across arbitrary decisions (Figures \@ref(fig:figure2)c and \@ref(fig:figure3)c), sample sizes (Figures \@ref(fig:figure2)d and \@ref(fig:figure3)d), and a specification grid showing the arbitrary data processing decisions associated with each effect (Figures \@ref(fig:figure2)e and \@ref(fig:figure3)e).

We preregistered all primary analyses on the Open Science Framework (OSF; see <https://osf.io/6r95z/>, including an updated primary analysis plan (see <https://osf.io/4vsnz/>), which included more details. We conducted secondary analyses after our primary analyses (see [Secondary Analyses](#secondary) below). Before conducting these analyses, we uploaded secondary analysis plans (see <https://osf.io/7fu35/> for the first plan and <https://osf.io/wcauf/> for the second plan), explaining the goals and purpose of each secondary analysis. A Red Team Critic [@redteammarket2021] successfully reproduced and verified (see <https://osf.io/xv94z/> for a report) all analyses prior to submission. All associated data and code are included in our preregistration repository on GitHub (<https://github.com/ethan-young/hidden-talents-multiverse>) and the OSF (<https://osf.io/6r95z/>), including source code to reproduce the manuscript itself.

@@

Overall, there were no significant interaction effects between task version and unpredictability, violence exposure, or poverty exposure (see Table \@ref(tab:table2), Table \@ref(tab:table3), and Figure \@ref(fig:figure2)b). However, analyses produced main effects of unpredictability (median $\beta$ = `r mod_terms2$shifting.unp$median_beta_main`, `r mod_terms2$shifting.unp$p_percent_main %>% str_remove("\\.00")` of *ps* \< .05, overall *p* = `r mod_terms2$shifting.unp$p_overall_main`) and violence exposure (median $\beta$ = `r mod_terms2$shifting.vio$median_beta_main`, `r mod_terms2$shifting.vio$p_percent_main` of *ps* \< .05, overall *p* = `r mod_terms2$shifting.vio$p_overall_main`; see Table \@ref(tab:table2)). There were no main effects of poverty exposure (median $\beta$ = `r mod_terms2$shifting.ses$median_beta_main`, no *ps* \< .05, overall *p* = `r mod_terms2$shifting.ses$p_overall_main`; see Table \@ref(tab:table2)). That is, exposure to more unpredictability and violence was associated with smaller switch costs, suggesting such exposures were associated with somewhat better shifting performance on average, irrespective of whether stimuli were abstract or ecologically relevant.

@@

Across each set of multiverse analyses, there were again significant main effects of task version. All median effects sizes for task version were $\beta$s = `r c(mod_terms1$updating.unp.type_z$median_beta, mod_terms1$updating.vio.type_z$median_beta, mod_terms1$updating.ses.type_z$median_beta) %>% mean()` (all *ps* \< .05) for models testing unpredictability, violence exposure, and poverty exposure, indicating that updating performance was better when stimuli were ecological relevant compared to abstract, on average (see SI Figures 4-6). There were also consistent main effects of age; median effects were $\beta$ = `r mod_terms1$updating.unp.age$median_beta` (all *ps* \< .05) for unpredictability models, $\beta$ = `r mod_terms1$updating.vio.age$median_beta` (all *ps* \< .05) for violence exposure models, and $\beta$ = `r mod_terms1$updating.ses.age$median_beta` (all *ps* \< .05) for poverty exposure models (see SI Figures 4-6). That is, older youth scored higher on working memory updating, on average. Finally, there were some main effects of test environment distraction ratings. Median effects sizes were $\beta$ = `r mod_terms1$updating.unp.test_envr$median_beta` (`r (mod_terms1$updating.unp.test_envr$p_percent *100) %>% round(2)`% of *ps* \< .05) for unpredictability models, $\beta$ = `r mod_terms1$updating.vio.test_envr$median_beta` (`r (mod_terms1$updating.vio.test_envr$p_percent *100) %>% round(2)`% of *ps* \< .05) for violence exposure models, and $\beta$ = `r mod_terms1$updating.ses.test_envr$median_beta` (`r (mod_terms1$updating.ses.test_envr$p_percent *100) %>% round(2)`% of *ps* \< .05) for poverty exposure models, indicating that more test distractions reduced working memory updating performance, on average (see SI Figures 4-6).

Multiverse analyses revealed a main effect of poverty exposure (median $\beta$ = `r mod_terms2$updating.ses$median_beta_main`, all *ps* \< .05, bootstrapped *p* \< .001) but no main effects of unpredictability (median $\beta$ = `r mod_terms2$updating.unp$median_beta_main`, no *ps* \< .05, bootstrapped *p* = `r mod_terms2$updating.unp$p_overall_main`) or violence exposure (median $\beta$ = `r mod_terms2$updating.vio$median_beta_main`, `r mod_terms2$updating.vio$p_percent_main` of *ps* \< .05, bootstrapped *p* = `r mod_terms2$updating.vio$p_overall_main`). Analyses also revealed two significant interactions: task version × violence exposure (median $\beta$ = `r mod_terms2$updating.vio$median_beta_int`, `r mod_terms2$updating.vio$p_percent_int` of *ps* \< .05, bootstrapped *p* = `r mod_terms2$updating.vio$p_overall_int`) and task version × poverty exposure (median $\beta$ = `r mod_terms2$updating.ses$median_beta_int`, `r mod_terms2$updating.ses$p_percent_int` of *ps* \< .05, bootstrapped *p* = `r mod_terms2$updating.ses$p_overall_int`). Youth exposed to low levels of violence or poverty tended to perform better on the abstract working memory updating task (violence exposure median *b~abstract~* = `r simple_effects$updating.vio$beta_abs`, `r simple_effects$updating.vio$p_percent_abs` of *ps* \< .05; poverty exposure median *b~abstract~* = `r simple_effects$updating.ses$beta_abs`, all *ps* \< .05), compared to youth exposed to high levels of violence and poverty (see Table \@ref(tab:table3) and Figure \@ref(fig:figure3)). However, this performance gap closed on the ecological working memory task (violence exposure median *b~ecological~* = `r simple_effects$updating.vio$beta_eco`, no *ps* \< .05; poverty exposure median *b~ecological~* = `r simple_effects$updating.ses$beta_eco`, `r simple_effects$updating.ses$p_percent_eco` of *ps* \< .05). Put differently, youth exposed to high levels of violence or poverty ($+$ 1 *SD*) performed better on the ecologically relevant version compared to their own performance on the abstract version (median simple *b~high-violence~* = `r simple_effects$updating.vio$beta_high`, all *ps* \< .05; median simple *b~high-poverty~* = `r simple_effects$updating.ses$beta_high`, all *ps* \< .05), whereas youth exposed to low violence or poverty ($-$ 1 *SD*) performed similarly across task versions (median simple *b~low-violence~* = `r simple_effects$updating.vio$beta_low`, no *ps* \< .05; median simple *b~low-poverty~* = `r simple_effects$updating.ses$beta_low`, no *ps* \< .05). In this sense, ecologically relevant stimuli appeared to equalize working memory updating performance among youth exposed to high levels of violence and poverty (see Table \@ref(tab:table3) and Figure \@ref(fig:figure3)a).

@@

The second question focused on the interactive effects of task version × violence exposure and task version × poverty exposure on working memory updating performance. Specifically, primary analyses revealed an equalization effect for both violence and poverty exposures on working memory updating. However, because measures of violence and poverty were not included in the same statistical model, it is unclear whether these interactions are independent. For example, one possibility is that exposures to violence and poverty are independently associated with working memory updating performance. Alternatively, they may overlap, reflecting the same association. To test this question, we ran a focused multiverse analysis with the same controls, main effects of task version, poverty and violence exposure, and two interaction terms: task-version × violence exposure and task-version × poverty in the same model. If violence and poverty exposures interact independently with task version, both interaction terms should remain significant. However, if they represent overlapping effects on working memory updating performance, one or the other may drop out of significance, which would suggest that one or the other adversity dimension is driving the effects of the other.

@@

For this expanded multiverse approach, we retained all original data processing decisions (i.e., original 64 combinations of decisions) and added adversity component as an additional decision. For unpredictability, we analyzed each unpredictability component and the composite measure (composite analyses are redundant with primary analyses but were retained for comparison purposes), which resulted in 384 analyses (64 original × three versions of unpredictability × two performance measures: attention shifting and working memory). We applied the same approach to violence exposure (three violence variables, 384 analyses) and poverty exposure (four poverty variables, 512 analyses). The full results of these analyses are reported in the supplement (see SI Table 2).

@@

Violence and poverty exposure interaction patterns were qualitatively similar to those obtained in our primary analyses (see Figure \@ref(fig:figure3)a and SI Figure 13). However, these secondary results suggest that the violence exposure interaction with task-version may be driving our poverty × task version results for working memory updating. One possibility is that the shared variance between poverty and violence exposures (*r* = `r txt5_cor$ses_vio$r`) is responsible for both interaction effects. In other words, although poverty and violence exposure are theoretically distinct, our poverty composite may capture poverty-related violence exposures (e.g., high poverty neighborhoods contain more crime) and therefore interact with task-version. When directly compared, explicit measures of violence exposure remained significant (although only under a small set of analytical decisions), whereas poverty interactions dropped out entirely.

@@

There are many possible explanations for the diverging adversity component patterns. First, some information sources may be more accurate than others. For example, asking about the number of fights one has been involved in or witnessed might be easier to report than asking broader questions about neighborhood danger. Second, for different cognitive abilities, objective versus subjective reports may be more relevant. For example, for attention-shifting, perceptions appeared to matter whereas objective events did for working memory updating. Third, although adversity components were correlated in the moderate range---as typically observed in the literature [@jacobs2012; @smith2021]---they may measure different constructs. For example, more objective indicators might capture aspects of the broader ecology but subjective, self-reported measures capture how youth have internalized and perceived their environment and family life. Although interesting, all secondary analyses are exploratory, and should be interpreted with caution. Nonetheless, these analyses highlight the need to measure multiple dimensions of adversity, compare them, and consider each measure's report format and source.

Taken together, our equalization patterns of results are striking compared to the backdrop of developmental science, which almost exclusively reports lowered cognitive performance in people exposed to harsh environments. We also document lowered performance among adversity-exposed youth, but this effect was specific to abstract stimuli, at least for working memory updating. When testing materials were more concrete, adversity-exposed youth perform about as well as youth from supportive backgrounds. These results converge with recent work comparing stimulus types among adversity-exposed youth. For example, Rifkin-Graboi and colleagues [-@rifkin-graboi2021] found preschoolers exposed to caregiver adversity were more accurate on relational memory for socioemotional than non-socioemotional stimuli and Zuilkowski and colleagues [-@zuilkowski2016] found that Sub-Saharan African children from low SES backgrounds were more accurate on reasoning tests with three dimensional stimuli than two dimensional stimuli.

@@

Theoretically, our findings support the notion that hidden talents may emerge when using ecologically relevant testing materials, but they also raise questions. In particular, why should ecologically relevant stimuli equalize working memory updating performance for adversity-exposed youth than youth with reduced exposures? There are at least three possibilities. First, adversity exposure could simultaneously lower and improve performance. While abstract working memory updating performance is lowered, ecological working memory updating performance is improved. However, this explanation requires that different cognitive mechanisms process different classes of stimuli, even on tests of the same ability. If this is true, those mechanisms that process abstract content underperform compared to those that process ecological content. Second, there may be no actual differences in underlying abilities. Instead, ecologically relevant stimuli may activate intact, but dormant or underrecruited, working memory updating processes. Thus, when adversity-exposed youth are tested under abstract testing conditions, their cognitive processes operate at a lower level (remain inactive) but operate equally well with ecological testing materials. This makes sense if adversity limits energy or cognitive resources. Under such constraints, it may be adaptive to activate expensive cognitive processes only when necessary. Third, ecological testing materials may remove barriers, such as testing anxiety or test unfamiliarity, that disadvantage adversity-exposed youth. Unlike the selective activation of cognitive resources, the additional stressors of testing are removed, allowing cognitive mechanisms to operate unimpeded. However, this explanation is difficult to reconcile with experimental work which finds that mild stress and uncertainty improves performance [@mittal2015; @young2018]. Future research should use research designs that can disentangle these explanations.

>>>>

\noindent Adversity-exposed youth tend to score lower on cognitive tests. However, the hidden talents approach proposes some abilities are enhanced by adversity, especially under ecologically relevant conditions. Two versions of an attention shifting and working memory updating task---one abstract, one ecological---were administered to `r txt1_sam$non_arb$n$total` youth (*M~age~* = `r txt1_sam$non_arb$age$mean`, *SD~age~* = `r txt1_sam$non_arb$age$sd`; `r (txt1_sam$non_arb$sex$females/txt1_sam$non_arb$n$total * 100) %>% round(2)`% female; `r ((txt1_sam$non_arb$n$total-txt1_sam$non_arb$race$non_white)/txt1_sam$non_arb$n$total * 100) %>% round(2)`% White). Measures of environmental unpredictability, violence, and poverty were collected to test adversity × task version interactions. There were no interactions for attention shifting. For working memory updating, youth exposed to violence and poverty scored lower than their peers with abstract stimuli but equally well with ecological stimuli. These results are striking compared to contemporary developmental science, which often reports lowered performance among adversity-exposed youth.

@@

One way to test this question is to replace sterile testing materials with ecologically relevant content from the real world. For example, one study tested memory and reasoning among young adults who varied in their exposures to violence [@frankenhuis2020a]. They created two versions of a memory and reasoning task, one neutral (chronological age) and one more ecologically relevant (social dominance). They found that people with more *current* exposures to violence were equally, or more, accurate when memorizing social-dominance relationships than people with fewer exposures. Notably, such violence exposures were associated with less accurate memory for neutral (age) relationships. This pattern was specific to *current* exposures: people with more *childhood* exposures to violence memorized both content types less accurately. There were no associations between violence exposures and reasoning performance. Overall, this study suggests that ecologically relevant testing materials may, in some conditions, 'level the playing field' for adversity-exposed youth [@ellis2020; @frankenhuis2020; @rifkin-graboi2021; @vantassel-baska2018].

Although research testing performance in ecologically relevant contexts or situations and using ecologically relevant materials have generated innovative ideas and findings, it remains unclear whether or how they can be practically applied. For example, narrow abilities, such as memorizing social dominance hierarchies or accurately detecting anger in faces, seem to have limited practical value in educational contexts. In addition, although executive functions and working memory are related to many educational outcomes, it is unclear how schools and teachers might leverage skills and abilities that specifically depend on inducing uncertainty. That seems impractical, at best, and unethical, at worst.

@@

Environmental unpredictability is difficult to measure and different definitions exist alongside each other [@young2020]. Despite this, we aligned our measurement approach with previous literature in two ways. First, we measured caregiver and residential instability, which are both thought to be proxies to environmental unpredictability, using interview-based measures [@ellis2009]. Second, we measured the extent to which youth perceive their family context as unpredictable using a questionnaire [@young2018]. Our goal was to aggregate these two information sources to arrive at a "best estimate" of environmental unpredictability.

@@

Our models for attention shifting and working memory updating differed in one respect. Our sample included a small subset of siblings (*n* = `r txt6_sib$n`, `r txt6_sib$families` families). All sibling groups were pairs of two (`r txt6_sib$sizes$two` sibling pairs) except for `r english::as.english(txt6_sib$sizes$three)` group of three and `r english::as.english(txt6_sib$sizes$four)` group of four siblings. We computed intraclass correlations for sibling groups for attention-shifting and working memory updating scores. Intraclass correlations for attention shifting were essentially zero. However, for working memory updating scores, intraclass correlations were `r txt6_sib$iccs$abs_update` and `r txt6_sib$iccs$eco_update` for the standard and ecological versions, respectively. Thus, all models analyzing working memory updating included an additional random intercept for siblings. Non-siblings were given the same sibling-group code, so the model could compute an intercept for non-siblings.

@@

Table \@ref(tab:table1) displays bivariate associations between all composite measures of adversity, including components of each, and descriptive statistics. Our primary interests were the main effect of adversity and the interaction of adversity with task-version. Table \@ref(tab:table2) shows median effect sizes ($\beta$-coefficients) and bootstrapped *p*-values for attention-shifting and working memory updating analyses for the main effect of adversity (unpredictability, violence, and poverty) and the interaction term (task version × adversity). Table \@ref(tab:table3) reports median simple effects of task version across high and low levels of adversity and median simple effects of adversity across abstract and ecologically relevant task versions.

@@

Figures \@ref(fig:figure2) and \@ref(fig:figure3) visualize multiverse task version × adversity interaction results for attention shifting and working memory updating performance, including depictions of performance on abstract and ecological task versions across high ($+$ 1 *SD*) and low ($-$ 1 *SD*) adversity exposure (Figures \@ref(fig:figure2)a and \@ref(fig:figure3)a), *p*-curves associated with each interaction term (Figures \@ref(fig:figure2)b and \@ref(fig:figure3)b), sorted interaction $\beta$-coefficients across arbitrary decisions (Figures \@ref(fig:figure2)c and \@ref(fig:figure3)c), sample sizes (Figures \@ref(fig:figure2)d and \@ref(fig:figure3)d), and a specification grid showing the arbitrary data processing decisions associated with each effect (Figures \@ref(fig:figure2)e and \@ref(fig:figure3)e).

We preregistered all primary analyses on the Open Science Framework (OSF; see <https://osf.io/6r95z/>), including an updated primary analysis plan (see <https://osf.io/4vsnz/>), which included more details. We conducted secondary analyses after our primary analyses (see [Secondary Analyses](#secondary) below). Before conducting these analyses, we uploaded secondary analysis plans (see <https://osf.io/7fu35/> for the first plan and <https://osf.io/wcauf/> for the second plan), explaining the goals and purpose of each secondary analysis. A Red Team Critic [@redteammarket2021] successfully reproduced and verified (see <https://osf.io/xv94z/> for a report) all analyses prior to submission. All associated data and code are included in our preregistration repository on GitHub (<https://github.com/ethan-young/hidden-talents-multiverse>) and the OSF (<https://osf.io/6r95z/>), including source code to reproduce the manuscript itself.

@@

Overall, there were no significant interaction effects between task version and unpredictability, violence exposure, or poverty exposure (see Table \@ref(tab:table2), Table \@ref(tab:table3), and Figure \@ref(fig:figure2)b). However, analyses produced main effects of unpredictability (median $\beta$ = `r mod_terms2$shifting.unp$median_beta_main`, `r mod_terms2$shifting.unp$p_percent_main %>% str_remove("\\.00")` of *ps* \< .05, overall *p* = `r mod_terms2$shifting.unp$p_overall_main`) and violence exposure (median $\beta$ = `r mod_terms2$shifting.vio$median_beta_main`, `r mod_terms2$shifting.vio$p_percent_main` of *ps* \< .05, overall *p* = `r mod_terms2$shifting.vio$p_overall_main`; see Table \@ref(tab:table2)). There were no main effects of poverty exposure (median $\beta$ = `r mod_terms2$shifting.ses$median_beta_main`, no *ps* \< .05, overall *p* = `r mod_terms2$shifting.ses$p_overall_main`; see Table \@ref(tab:table2)). That is, exposure to more unpredictability and violence was associated with smaller switch costs, suggesting such exposures were associated with somewhat better shifting performance on average, irrespective of whether stimuli were abstract or ecological.

@@

Across each set of multiverse analyses, there were again significant main effects of task version. All median effects sizes for task version were $\beta$s = `r c(mod_terms1$updating.unp.type_z$median_beta, mod_terms1$updating.vio.type_z$median_beta, mod_terms1$updating.ses.type_z$median_beta) %>% mean()` (all *ps* \< .05) for models testing unpredictability, violence exposure, and poverty exposure, indicating that updating performance was better when stimuli were ecological compared to abstract, on average (see SI Figures 4-6). There were also consistent main effects of age; median effects were $\beta$ = `r mod_terms1$updating.unp.age$median_beta` (all *ps* \< .05) for unpredictability models, $\beta$ = `r mod_terms1$updating.vio.age$median_beta` (all *ps* \< .05) for violence exposure models, and $\beta$ = `r mod_terms1$updating.ses.age$median_beta` (all *ps* \< .05) for poverty exposure models (see SI Figures 4-6). That is, older youth scored higher on working memory updating, on average. Finally, there were some main effects of test environment distraction ratings. Median effects sizes were $\beta$ = `r mod_terms1$updating.unp.test_envr$median_beta` (`r (mod_terms1$updating.unp.test_envr$p_percent *100) %>% round(2)`% of *ps* \< .05) for unpredictability models, $\beta$ = `r mod_terms1$updating.vio.test_envr$median_beta` (`r (mod_terms1$updating.vio.test_envr$p_percent *100) %>% round(2)`% of *ps* \< .05) for violence exposure models, and $\beta$ = `r mod_terms1$updating.ses.test_envr$median_beta` (`r (mod_terms1$updating.ses.test_envr$p_percent *100) %>% round(2)`% of *ps* \< .05) for poverty exposure models, indicating that more test distractions reduced working memory updating performance, on average (see SI Figures 4-6).

Multiverse analyses revealed a main effect of poverty exposure (median $\beta$ = `r mod_terms2$updating.ses$median_beta_main`, all *ps* \< .05, bootstrapped *p* \< .001) but no main effects of unpredictability (median $\beta$ = `r mod_terms2$updating.unp$median_beta_main`, no *ps* \< .05, bootstrapped *p* = `r mod_terms2$updating.unp$p_overall_main`) or violence exposure (median $\beta$ = `r mod_terms2$updating.vio$median_beta_main`, `r mod_terms2$updating.vio$p_percent_main` of *ps* \< .05, bootstrapped *p* = `r mod_terms2$updating.vio$p_overall_main`). Analyses also revealed two significant interactions: task version × violence exposure (median $\beta$ = `r mod_terms2$updating.vio$median_beta_int`, `r mod_terms2$updating.vio$p_percent_int` of *ps* \< .05, bootstrapped *p* = `r mod_terms2$updating.vio$p_overall_int`) and task version × poverty exposure (median $\beta$ = `r mod_terms2$updating.ses$median_beta_int`, `r mod_terms2$updating.ses$p_percent_int` of *ps* \< .05, bootstrapped *p* = `r mod_terms2$updating.ses$p_overall_int`). Youth exposed to low levels of violence or poverty tended to perform better on the abstract working memory updating task (violence exposure median *b~abstract~* = `r simple_effects$updating.vio$beta_abs`, `r simple_effects$updating.vio$p_percent_abs` of *ps* \< .05; poverty exposure median *b~abstract~* = `r simple_effects$updating.ses$beta_abs`, all *ps* \< .05), compared to youth exposed to high levels of violence and poverty (see Table \@ref(tab:table3) and Figure \@ref(fig:figure3)). However, this performance gap closed on the ecological working memory task (violence exposure median *b~ecological~* = `r simple_effects$updating.vio$beta_eco`, no *ps* \< .05; poverty exposure median *b~ecological~* = `r simple_effects$updating.ses$beta_eco`, `r simple_effects$updating.ses$p_percent_eco` of *ps* \< .05). Put differently, youth exposed to high levels of violence or poverty ($+$ 1 *SD*) performed better on the ecological version compared to their own performance on the abstract version (median simple *b~high-violence~* = `r simple_effects$updating.vio$beta_high`, all *ps* \< .05; median simple *b~high-poverty~* = `r simple_effects$updating.ses$beta_high`, all *ps* \< .05), whereas youth exposed to low violence or poverty ($-$ 1 *SD*) performed similarly across task versions (median simple *b~low-violence~* = `r simple_effects$updating.vio$beta_low`, no *ps* \< .05; median simple *b~low-poverty~* = `r simple_effects$updating.ses$beta_low`, no *ps* \< .05). In this sense, ecological stimuli appeared to equalize working memory updating performance among youth exposed to high levels of violence and poverty (see Table \@ref(tab:table3) and Figure \@ref(fig:figure3)a).

@@

The second question focused on the interactive effects of task version × violence exposure and task version × poverty exposure on working memory updating performance. Specifically, primary analyses revealed an equalization effect for both violence and poverty exposures on working memory updating. However, because measures of violence and poverty were not included in the same statistical model, it is unclear whether these interactions are independent. For example, one possibility is that exposures to violence and poverty are independently associated with working memory updating performance. Alternatively, they may overlap, reflecting the same association. To test this question, we ran a focused multiverse analysis with the same controls, main effects of task version, poverty and violence exposure, and two interaction terms: task-version × violence exposure and task-version × poverty exposure in the same model. If violence and poverty exposures interact independently with task version, both interaction terms should remain significant. However, if they represent overlapping effects on working memory updating performance, one or the other may drop out of significance, which would suggest that one or the other adversity dimension is driving the effects of the other.

@@

For this expanded multiverse approach, we retained all original data processing decisions (i.e., original 64 combinations of decisions) and added adversity component as an additional decision. For unpredictability, we analyzed each unpredictability component and the composite measure (composite analyses are redundant with primary analyses but were retained for comparison purposes), which resulted in 384 analyses (64 original × three versions of unpredictability × two performance measures: attention shifting and working memory updating). We applied the same approach to violence exposure (three violence variables, 384 analyses) and poverty exposure (four poverty variables, 512 analyses). The full results of these analyses are reported in the supplement (see SI Table 2).

@@

Violence and poverty exposure interaction patterns were qualitatively similar to those obtained in our primary analyses (see Figure \@ref(fig:figure3)a and SI Figure 13). However, these secondary results suggest that the violence exposure interaction with task version may be driving our poverty × task version results for working memory updating. One possibility is that the shared variance between poverty and violence exposures (*r* = `r txt5_cor$ses_vio$r`) is responsible for both interaction effects. In other words, although poverty and violence exposure are theoretically distinct, our poverty composite may capture poverty-related violence exposures (e.g., high poverty neighborhoods contain more crime) and therefore interact with task version. When directly compared, explicit measures of violence exposure remained significant (although only under a small set of analytical decisions), whereas poverty interactions dropped out entirely.

@@

There are many possible explanations for the diverging adversity component patterns. First, some information sources may be more accurate than others. For example, asking about the number of fights one has been involved in or witnessed might be easier to report than asking broader questions about neighborhood danger. Second, for different cognitive abilities, objective versus subjective reports may be more relevant. For example, for attention shifting, perceptions appeared to matter whereas objective events did for working memory updating. Third, although adversity components were correlated in the moderate range---as typically observed in the literature [@jacobs2012; @smith2021]---they may measure different constructs. For example, more objective indicators might capture aspects of the broader ecology but subjective, self-reported measures capture how youth have internalized and perceived their environment and family life. Although interesting, all secondary analyses are exploratory, and should be interpreted with caution. Nonetheless, these analyses highlight the need to measure multiple dimensions of adversity, compare them, and consider each measure's report format and source.

Taken together, our equalization patterns of results are striking compared to the backdrop of developmental science, which almost exclusively reports lowered cognitive performance in people exposed to harsh environments. We also document lowered performance among adversity-exposed youth, but this effect was specific to abstract stimuli, at least for working memory updating. When testing materials were more concrete, adversity-exposed youth perform about as well as youth from supportive backgrounds. These results converge with recent work comparing stimulus types among adversity-exposed youth. For example, Rifkin-Graboi and colleagues [-@rifkin-graboi2021] found preschoolers exposed to caregiver adversity were more accurate on relational memory for socioemotional than non-socioemotional stimuli and Zuilkowski and colleagues [-@zuilkowski2016] found that Sub-Saharan African children from high poverty backgrounds were more accurate on reasoning tests with three dimensional stimuli than two dimensional stimuli.

@@

Theoretically, our findings support the notion that hidden talents may emerge when using ecologically relevant testing materials, but they also raise questions. In particular, why should ecologically relevant stimuli equalize working memory updating performance for adversity-exposed youth? There are at least three possibilities. First, adversity exposure could simultaneously lower and improve performance. While abstract working memory updating performance is lowered, ecological working memory updating performance is improved. However, this explanation requires that different cognitive mechanisms process different classes of stimuli, even on tests of the same ability. If this is true, those mechanisms that process abstract content underperform compared to those that process ecological content. Second, there may be no actual differences in underlying abilities. Instead, ecologically relevant stimuli may activate intact, but dormant or underrecruited, working memory updating processes. Thus, when adversity-exposed youth are tested under abstract testing conditions, their cognitive processes operate at a lower level (remain inactive) but operate equally well with ecological testing materials. This makes sense if adversity limits energy or cognitive resources. Under such constraints, it may be adaptive to activate expensive cognitive processes only when necessary. Third, ecological testing materials may remove barriers, such as testing anxiety or test unfamiliarity, that disadvantage adversity-exposed youth. Unlike the selective activation of cognitive resources, the additional stressors of testing are removed, allowing cognitive mechanisms to operate unimpeded. However, this explanation is difficult to reconcile with experimental work which finds that mild stress and uncertainty improves performance [@mittal2015; @young2018]. Future research should use methods and designs that can disentangle these explanations.
